|

### 2023-10-14 chapter_2/src/roberts_empirical_chapter_1.qmd
drafting some changes

Now that I have presented some background and have introduced the snap-judgment model of politically-relevant visual information, this chapter is a first attempt to tackle the question, "Do colors provide political information that shape attitudes?" There are two goals that I am hoping to achieve in this chapter. The first is to convince the reader that the colors red and blue convey affiliative information. The second is that these connections between colors and political groups can shape political outcomes.

This chapter applies the snap-judgment model to a simple case: the use of color on yard signs used by campaigns in elections. I focus on yard signs as they are a common form of campaign marketing and evidence suggests that they can shape individual behavior [@makse_politics_2019] and electoral outcomes [@green_effects_2016]. Additionally, I start by examining the colors red and blue. While there are other colors that may provide political information -- black and white for Black Lives Matter and Blue Lives Matter movements, the overarching goal of this book is to lay the foundation for a larger research agenda in visual political psychology [@bucy_editors_2021]. Without such foundations, we would be jumping into the study of complex cases and circumstances without a firm grasp of how things matter on a more fundamental level.

Before getting into things, it is worth noting that while there indeed is empirical and experiential evidence suggesting that Republicans use the color red and Democrats use the color blue [@schloss_politics_2014;@williams_whats_2022], there is currently no theoretical framework that explains how these connections happen at the individual level. As a side note, one might find it unsatisfying that I do not address *how* these associations between color and the parties came about. While the connections between Republicans with the color red and Democrats with the color blue seemed to quicken after the 2000 Presidential election (with news networks consistently using those colors in campaign coverage [@elving_color_2014]), it is not clear if this was a deliberate choice, or whether this was yet another symptom of an increasingly polarized public. An undertaking to address the specifics of this part of American political history would likely be a different book project.

>>>>

Do colors provide political information that shape attitudes among the mass public? There are two goals that I am hoping to achieve in this chapter. The first is to convince the reader that the colors red and blue convey affiliative information. The second is that these connections between colors and political groups can shape political outcomes.

Many understand that colors are influential to shaping attitudes and behaviors in a number of ways. Color theorists and scholars of marketing understand color as having important implications for reactions to a product and resulting consumer behavior. For example, @kuo_impact_2023 demonstrate that hotel rooms that use more "cool colors" (blues, greens, etc.) evoke positive affect for potential customers and that these positive affective responses are predictive of their eventual choice to book the room. Colors mean a varying number of things, however. As a gross, yet intelligible example, yellow sweaters relative to yellow bodily fluids can elicit strong and very different responses.

The role of visuals in politics is receiving more attention among interdisciplinary scholars in fields like communication, political science, and the psychological sciences. While many of these studies examine fascinating but complicated forms of visual information such as symbology [@williams_whats_2022] and visual framing [@grabe_image_2009], we have yet to integrate these interdisciplinary perspectives with existing models of political information processing to understand how these visual forms of information act as cues [see @lilleker_power_2019] but also what the consequences are for political behavior among the mass public.

Building upon descriptive evidence suggesting that Democratic campaigns use the color blue more than Republican campaigns [@williams_whats_2022] and that Republican voters express a preference for red during election season [@schloss_politics_2014], I present a snap-judgment model of politically-relevant visual information processing that outlines the cognitive processes that drive these associations and how those associations drive important political outcomes. I use theories of associative memory and spreading activation [see @collins_spreading-activation_1975] to conceptualize the association of the colors red and blue with Republicans and Democrats as two interconnected nodes. Informed by conceptualizations of these connections as carrying affective information informed by perspectives on biological survival reaching back to Darwin [@ralph_neuroscience_2018], I argue that when these nodes are activated by exposure to these colors in political contexts, other related nodes become active and contain affective information. As partisanship evokes strong affective reactions for many members of the public [@iyengar_affect_2012], this activation elicits an affective reaction. Considering that most visual information and that these processes of association occur pre-consciously [@ames_impression_2012;@fazio_2007;@sander_cambridge_2013], I argue that the elicited affective response to such information can be "transferred" [see @morris_activation_2003;@taber_illusion_2016] to later post-cognitive processing of more traditional forms of political information such as stated policy positions. The implication being that these associations can inform reactions to political information and thus political behavior subconsciously.
<!--
This chapter applies the snap-judgment model to a simple case: the use of color on yard signs used by campaigns in elections. I focus on yard signs as they are a common form of campaign marketing and evidence suggests that they can shape individual behavior [@makse_politics_2019] and electoral outcomes [@green_effects_2016]. Additionally, I start by examining the colors red and blue. While there are other colors that may provide political information -- black and white for Black Lives Matter and Blue Lives Matter movements, the overarching goal of this book is to lay the foundation for a larger research agenda in visual political psychology [@bucy_editors_2021]. Without such foundations, we would be jumping into the study of complex cases and circumstances without a firm grasp of how things matter on a more fundamental level.

Before getting into things, it is worth noting that while there indeed is empirical and experiential evidence suggesting that Republicans use the color red and Democrats use the color blue [@schloss_politics_2014;@williams_whats_2022], there is currently no theoretical framework that explains how these connections happen at the individual level. As a side note, one might find it unsatisfying that I do not address *how* these associations between color and the parties came about. While the connections between Republicans with the color red and Democrats with the color blue seemed to quicken after the 2000 Presidential election (with news networks consistently using those colors in campaign coverage [@elving_color_2014]), it is not clear if this was a deliberate choice, or whether this was yet another symptom of an increasingly polarized public. An undertaking to address the specifics of this part of American political history would likely be a different book project.
-->
